---
title: "group_document"
format:
  html:
    embed-resources: true
editor: visual
---

# Lab_07 Assignment Group 12

*Group members:*

s22277, s223712, s234118, 222940, 222785

## PCA

For this assignment we want to learn to do a PCA (principal component analysis) in a collaboration by the group - facilitated by the use of github. For this purpose, the workflow of Professor Claus O. Wilke was followed.

## Load libraries

```{r}
#| echo: true
#| eval: true
library(tidyverse)
library(broom)
```

## Load data

First we load the data - we chose to use the data from "PCA tidyverse style" by Claus O. Wilke, Professor of Integrative Biology (https://clauswilke.com/blog/2020/09/07/pca-tidyverse-style/).

This data is a breast cancer dataset from the University of Wisconsin Hospitals. 699 patient biopsies are available, and is scored on a scale of 1-10 for 9 attributes. The true outcome is binary (benign/malignant).

```{r}
biopsy <- read_csv("https://wilkelab.org/classes/SDS348/data_sets/biopsy.csv")
biopsy |> 
  head()
```

## First we look at the data in PC coordinates

```{r}
pca_fit <- biopsy |> 
  select(where(is.numeric)) |> 
  scale() |> 
  prcomp()
pca_fit |> 
  augment(biopsy) |> 
  ggplot(mapping = aes( x = .fittedPC1, 
                        y = .fittedPC2, 
                        color = outcome)) + 
  geom_point() +
  scale_color_manual(
    values = c(malignant = "red", benign = "blue")) +
  theme_minimal() +
  labs(x = "PC1",
       y = "PC2",
       title = "PCA",
       caption = "Data from https://clauswilke.com/blog/2020/09/07/pca-tidyverse-style/") +
  theme(plot.caption = element_text(hjust = 1))
```

After reviewing the PCA plot, it becomes evident that the malignant and benign data points distinctly separate when projected onto the first two principal components. This results in two overlapping yet clearly distinguishable clusters (red one for malignant and blue for benign).

## We now plot the rotation matrix

```{r}
pca_fit |> 
  tidy(matrix = "rotation") |> 
  pivot_wider(names_from = "PC", 
              names_prefix = "PC",
              values_from = "value") |> 
  ggplot(mapping = aes(x = PC1, 
                       y = PC2))+
  geom_segment(xend = 0, 
               yend = 0) +
  geom_text(
    aes(label = column),
    hjust = 1, 
    vjust = 1,
    size = 3,
    color = "#098a7f") +
  theme_minimal() +
  xlim(-1.25, .5) + 
  ylim(-.5, 1) + 
  labs(title = "Rotation Matrix",
       caption = "Data from https://clauswilke.com/blog/2020/09/07/pca-tidyverse-style/") +
  theme(plot.caption = element_text(hjust = 0)) +
  coord_fixed()
```

## We quantify the variance explained by each PC

```{r}
pca_fit |>
  tidy(matrix = "eigenvalues")|> 
  ggplot(mapping = 
           aes(x = PC, 
               y = percent)) +
  geom_col(fill = "#abd94e", alpha = 0.5) +
  scale_x_continuous(breaks = 1:9) +
  scale_y_continuous(
    labels = scales::percent_format(),
    expand = expansion(mult = c(0, 0.01))
  ) +
  labs(title = "Variance captured by each PC", 
       caption = "Data from https://clauswilke.com/blog/2020/09/07/pca-tidyverse-style/")
  theme_classic()
```

Each PC captures a certain amount of the total variance in the dataset. The first component captures 65% of the variation which presents high importance for the data, separating the benign samples from the malignant samples.
